---
title: "Grouped Hyper Data Frame"
author: Tingting Zhan
date: "`r format(Sys.time(), 'Last updated %d %B, %Y')`"
format: 
  html:
    page-layout: full
    html-math-method: katex
toc: true
toc-location: left
toc-depth: 4
toc-title: ''
editor: visual
bibliography: groupedHyperframe.bib
knitr:
  opts_chunk: 
    collapse: true
    comment: "#>" 
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette of R package **`groupedHyperframe`** ([CRAN](https://cran.r-project.org/package=groupedHyperframe), [Github](https://github.com/tingtingzhan/groupedHyperframe), [RPubs](https://rpubs.com/tingtingzhan/groupedHyperframe)) documents

-   the creation of `groupedHyperframe` object,
-   the batch processes for a `hyperframe` and/or `groupedHyperframe`, and
-   the aggregation of various statistics over (nested) grouping structure.

R terminology might be different from that of mathematics and statistics.  Please refer to Appendix @sec-terms for explanation and reference.

## Prerequisite

Package **`groupedHyperframe`** requires R version 4.5.0 (released 2025-04-11) or higher. [RStudio](https://posit.co/download/rstudio-desktop/) is not required, but highly recommended. This vignette is generated under `r R.version$version.string` using packages **`knitr`** [@knitr], **`quarto`** [@quarto], **`rmarkdown`** [@rmarkdown], and the development versions of the **`spatstat`** family of packages [@spatstat15; @spatstat05].

```{r}
#| label: prerequisite
#| warning: false
#| eval: false
remotes::install_github('spatstat/spatstat', upgrade = 'always')
remotes::install_github('spatstat/spatstat.data', upgrade = 'always')
remotes::install_github('spatstat/spatstat.explore', upgrade = 'always')
remotes::install_github('spatstat/spatstat.geom', upgrade = 'always')
remotes::install_github('spatstat/spatstat.linnet', upgrade = 'always')
remotes::install_github('spatstat/spatstat.model', upgrade = 'always')
remotes::install_github('spatstat/spatstat.random', upgrade = 'always')
remotes::install_github('spatstat/spatstat.sparse', upgrade = 'always')
remotes::install_github('spatstat/spatstat.univar', upgrade = 'always')
remotes::install_github('spatstat/spatstat.utils', upgrade = 'always')
```

Experimental (and maybe unstable) features are released *extremely frequently* to [Github](https://github.com/tingtingzhan/groupedHyperframe). [Active developers should use the Github version; suggestions and bug reports are welcome!]{style="background-color: #FFFF00"}

```{r}
#| warning: false
#| eval: false
remotes::install_github('tingtingzhan/groupedHyperframe')
```

Stable releases to [CRAN](https://CRAN.R-project.org/package=groupedHyperframe) are typically updated every 2 to 3 months, or when the authors have an upcoming manuscript in the peer-reviewing process.

```{r}
#| warning: false
#| eval: false
utils::install.packages('groupedHyperframe') # Developers, do NOT use!!
```

### Dependencies

Package **`groupedHyperframe`** `Imports` packages

-   **`cli`** [@cli, version `r packageVersion('cli')`], for attractive command line interfaces (CLIs)
-   `r knitr::asis_output('\U1f5dd')` **`matrixStats`** [@matrixStats, version `r packageVersion('matrixStats')`], key dependency, for matrix arithmetic
-   **`parallel`** (shipped with `r R.version$version.string`), for parallel computing
-   `r knitr::asis_output('\U1f5dd')` **`pracma`** [@pracma, version `r packageVersion('pracma')`], key dependency, for (cumulative) trapezoidal integration
-   `r knitr::asis_output('\U1f5dd')` **`spatstat.explore`** (developer version `r packageVersion('spatstat.explore')`) and **`spatstat.geom`** (developer version `r packageVersion('spatstat.geom')`), @spatstat15, key dependency, for spatial statistics
-   `r knitr::asis_output('\U1f5dd')` **`SpatialPack`** [@SpatialPack, version `r packageVersion('SpatialPack')`], key dependency, for Tjøstheim's coefficient of spatial association

## Getting Started

Examples in this vignette require that the `search` path has

```{r}
#| message: false
library(groupedHyperframe)
library(survival) # to help hyperframe understand Surv object
```

```{r}
#| echo: false
op = par(no.readonly = TRUE)
#options(mc.cores = 1L) # for CRAN submission
```

## Acknowledgement

This work is supported by National Institutes of Health, U.S. Department of Health and Human Services grants

-   R01CA222847 ([I. Chervoneva](https://orcid.org/0000-0002-9104-4505), [T. Zhan](https://orcid.org/0000-0001-9971-4844), and [H. Rui](https://orcid.org/0000-0002-8778-261X))
-   R01CA253977 (H. Rui and I. Chervoneva).

# Grouped Hyper Data Frame

An `S3` class `hyperframe`, for **hyper data frame**, is defined in package **`spatstat.geom`** [@spatstat15; @spatstat05]. A `hyperframe` object contains columns either as `vector`s like in a `data.frame`, or as `list`s of objects of the same class, a.k.a., the `hypercolumns`. The `hyperframe` data structure is particularly useful in spatial analysis, e.g., with medical images, where the spatial information in each image would be represented by one element in a `hypercolumn`.

```{r}
spatstat.data::demohyper
```

We introduce a new `S3` class `groupedHyperframe`, for **grouped hyper data frame**, which `inherits` from the `hyperframe` class. The derived class `groupedHyperframe` has additional `attributes`

-   `attr(., 'group')`, a one-sided `formula` of the (nested) grouping structure, e.g., `~patient/image` when each patient has one or more images. The grammar of the nested grouping structure $g_1/.../g_m$ (`~g1/.../gm`) follows that of the parameter `random` of functions `nlme::lme()` and `nlme::nlme()`. In fact, the `'grouped'` extension of a `hyperframe` is inspired by the `nlme::groupedData` class which inherits from `data.frame` [@nlme]. Appendix @sec-nestedGroup outlines the experimental features of nested grouping structure that are subject to change.

The `S3` generic function `as.groupedHyperframe()` converts objects of various classes into a `groupedHyperframe`. We also introduce aggregation functions `aggregate_*()` of the hypercolumns in a `groupedHyperframe`, at either one of the nested grouping levels $g_1,\cdots,g_{m-1}$. Aggregation at the lowest grouping level $g_m$ is ignored, i.e., no aggregation to be performed. Available aggregation methods are the point-wise minima `base::pmin()`, maxima `base::pmax()`, means `pmean()` (default) and medians `pmedian()`.

## From `data.frame`

The `S3` method dispatch `as.groupedHyperframe.data.frame()` converts a `data.frame` with substantial amount of duplicated information into a `groupedHyperframe`. This function

1.  inspects the input `data.frame` by the user-specified (nested) `group`ing structure;
2.  identifies the column(s) with ***non-identical elements*** within the lowest group, and converts them into hypercolumn(s);
3.  returns a `groupedHyperframe`.

In the following example, consider a toy data set **`wrobel_lung0`** with non-identical column *`hladr`* in the lowest group *`image_id`* of the nested grouping structure *`~patient_id/image_id`*.

```{r}
wrobel_lung0 = wrobel_lung |>
  within.data.frame(expr = {
    x = y = NULL
    dapi = phenotype = tissue = NULL
  })
```

```{r}
wrobel_lung0 |> head()
```

By converting **`wrobel_lung0`** into a `groupedHyperframe`, the numeric *`hladr`* from each *`~patient_id/image_id`* are converted into elements of the numeric-hypercolumn *`hladr`* in the returned object **`wrobel_lung0g`**. Each row of a `groupedHyperframe` represents the lowest group of the nested grouping structure. The R console output (`S3` method dispatch `print.groupedHyperframe()`) highlights the nested grouping structure, number of clusters at each grouping level, as well as the first 10 (or less) rows of the returned `groupedHyperframe`.

```{r}
(wrobel_lung0g = wrobel_lung0 |> as.groupedHyperframe(group = ~ patient_id/image_id))
```

Converting a `data.frame` with substantial amount of duplicated information into a `groupedHyperframe` greatly reduces the memory allocation. A `groupedHyperframe`, however, would not reduce much the `save`d `file.size` compared to a `data.frame`, if `xz` compression is used for both.

```{r}
#| code-fold: true
#| code-summary: "See for yourself"
unclass(object.size(wrobel_lung0g)) / unclass(object.size(wrobel_lung0)) # great reduction
f_g = tempfile(fileext = '.rds')
wrobel_lung0g |> saveRDS(file = f_g, compress = 'xz')
f = tempfile(fileext = '.rds')
wrobel_lung0 |> saveRDS(file = f, compress = 'xz')
file.size(f_g) / file.size(f) # not much reduction
```

### Aggregation of numeric-hypercolumn

Function `aggregate_quantile()` aggregates the `quantile`s of each element in the numeric-hypercolumn, e.g., *`hladr`* in **`wrobel_lung0g`**, by point-wise means (default of parameter `f_aggr_`) at the biologically independent group, e.g., *`~patient_id`*. The returned object is a `hyperframe` instead of a `groupedHyperframe`, as we have ***one*** aggregated *`hladr.quantile`* per *`~patient_id`*, thus eliminates the need for a grouping structure. The R console output is taken care of by function `spatstat.geom::print.hyperframe()`.

```{r}
#| message: false
wrobel_lung0g |>
  aggregate_quantile(by = ~ patient_id, probs = seq.int(from = .01, to = .99, by = .01))
```

## From `hyperframe`

The `S3` method dispatch `as.groupedHyperframe.hyperframe()` converts a `hyperframe`, e.g., from package **`spatstat.data`**, into a `groupedHyperframe`. This function simply inspects and adds a (nested) grouping structure to the input `hyperframe`.

In the following example, we inspect the data set `spatstat.data::osteo`, which has the serial number of sampling volume `brick` nested in the bone sample `id`, and add the nested grouping structure `~id/brick` to it.

```{r}
spatstat.data::osteo |> 
  as.groupedHyperframe(group = ~ id/brick)
```

In this vignette, we do not place much emphasize on the objects provided in package **`spatstat.data`**, for now.

# Grouping `ppp`-Hypercolumn

In this section, we first introduce the creation of a `groupedHyperframe` with *one-and-only-one* point pattern (`ppp`) hypercolumn. Then, we discuss

-   the batch processes of spatial point pattern analyses on the *one-and-only-one* `ppp`-hypercolumn of a `hyperframe` (and/or `groupedHyperframe`). These batch processes are not intended for a `hyperframe` (and/or `groupedHyperframe`) with multiple `ppp`-hypercolumn in the foreseeable future, as that would require checking for name clashes in the `marks` from multiple `ppp`-hypercolumn.
-   the aggregation of batch process returns over the nested grouping structure.

Function `grouped_ppp()` creates a `groupedHyperframe` with *one-and-only-one* `ppp`-hypercolumn. In the following example, the argument `formula` specifies

-   the point pattern `marks`, e.g., numeric mark *`hladr`* and `multitype` mark *`phenotype`*, on the left-hand-side
-   the additional predictors and/or endpoints for downstream analysis, e.g., *`OS`*, *`gender`* and *`age`*, before the `|` separator on the right-hand-side
-   the (nested) grouping structure, e.g., *`image_id`* nested in *`patient_id`*, after the `|` separator on the right-hand-side.

```{r}
(s = wrobel_lung |>
   grouped_ppp(formula = hladr + phenotype ~ OS + gender + age | patient_id/image_id))
```

## Batch processes {#sec-batch}

### Return of `fv`-hypercolumn

In this section, we discuss the batch processes that return a function value table (`fv`) hypercolumn, i.e., a hypercolumn which consists of a `list` of `fv.object`s.

| Batch Process | Workhorse in package **`spatstat.explore`** | `fv`-hypercolumns Suffix |
|------------------------|------------------------|------------------------|
| `Emark_()` and `Vmark_()` | [`Emark` and `Vmark`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/Emark.html), conditional mean $E(r)$ and variance $V(r)$, diagnostics for dependence between the points and the marks [@Emark] | `.E` and `.V` |
| `markcorr_()` | [`markcorr`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/markcorr.html), marked correlation $k_{mm}(r)$ or generalized mark correlation $k_f(r)$ [@markcorr] | `.k` |
| `markvario_()` | [`markvario`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/markvario.html), mark variogram $\gamma(r)$ [@markvario] | `.gamma` |
| `Kmark_()` | [`Kmark`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/Kmark.html), mark-weighted $K_f(r)$ function [@Kmark] | `.K` |

: Batch processes applicable to numeric `marks`

| Batch Process | Workhorse in package **`spatstat.explore`** | `fv`-hypercolumns Suffix |
|------------------------|------------------------|------------------------|
| `Gcross_()` | [`Gcross`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/Gcross.html), multitype nearest-neighbour distance $G_{ij}(r)$ | `.G` |
| `Kcross_()` | [`Kcross`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/Kcross.html), multitype $K_{ij}(r)$ | `.K` |
| `Jcross_()` | [`Jcross`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/Jcross.html), multitype $J_{ij}(r)$ [@Jcross] | `.J` |
| `Lcross_()` | [`Lcross`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/Lcross.html), multitype $L_{ij}(r)=\sqrt{\frac{K_{ij}(r)}{\pi}}$ | `.L` |

: Batch processes applicable to `multitype` `marks`

### Return of numeric-hypercolumn

In this section, we discuss the batch processes that return a numeric hypercolumn, i.e., a hypercolumn which consists of a `list` of numeric `vector`s.

| Batch Process | Workhorse in package **`spatstat.geom`** | Applicable to | numeric-hypercolumns Suffix |
|------------------|------------------|------------------|------------------|
| `nncross_()` | [`nncross.ppp`](https://search.r-project.org/CRAN/refmans/spatstat.geom/html/nncross.html)`(., what = 'dist')`, nearest neighbour distance | `multitype` marks | `.nncross` |

### In a pipeline

Multiple batch processes may be applied to a `hyperframe` (and/or `groupedHyperframe`) in a pipeline using the native pipe operator `|>` introduced since R version 4.1.0.

```{r}
r = seq.int(from = 0, to = 250, by = 10)
out = s |>
  Emark_(r = r, correction = 'none') |> # slow
  # Vmark_(r = r, correction = 'none') |> # slow
  # markcorr_(r = r, correction = 'none') |> # slow
  # markvario_(r = r, correction = 'none') |> # slow
  # Kmark_(r = r, correction = 'none') |> # fast
  Gcross_(i = 'CK+.CD8-', j = 'CK-.CD8+', r = r, correction = 'none') |> # fast
  # Kcross_(i = 'CK+.CD8-', j = 'CK-.CD8+', r = r, correction = 'none') |> # fast
  nncross_(i = 'CK+.CD8-', j = 'CK-.CD8+', correction = 'none') # fast
```

The `message` of "Legal $r_\text{max}$" provides the indices in the `ppp`-hypercolumn, i.e., the row indices of the `hyperframe`, where the legal $r_\text{max}$ (see Appendix @sec-legal_rmax) is smalller than the user-specified $r_\text{max}$. The returned `hyperframe` (and/or `groupedHyperframe`) has

-   `fv`-hypercolumn *`hladr.E`*, created by function `Emark_()` on numeric mark *`hladr`*
-   `fv`-hypercolumn *`phenotype.G`*, created by function `Gcross_()` on `multitype` mark *`phenotype`*
-   numeric-hypercolumn *`phenotype.nncross`*, created by function `nncross_()` on `multitype` mark *`phenotype`*

```{r}
out
```

## Truncated Aggregation

In this section, we discuss the aggregation over nested grouping structure.

### Of `fv`-hypercolumn(s)

Function `aggregate_fv()` performs (truncated) aggregation of

-   the *function values*, i.e., the black-solid-curve of `plot.fv`.
-   the *cumulative trapezoidal integration* under the black-solid-curve.

When a user-specified $r_\text{max}$ is provided to a batch process (in @sec-batch) on *all* `ppp.object`s in the `ppp`-hypercolumn, inevitably some of the `fv`-returns may contain exceptional/illegal function values (see Appendix @sec-legal_rmax). In this case, the aggregation is *truncated* at the minimum legal $r_\text{max}$ of the `fv.object`s in each `fv`-hypercolumn. In the following example, we have

-   numeric-hypercolumns *`hladr.E.value`* and *`hladr.E.cumtrapz`*, the aggregated function values and aggregated cumulative trapezoidal integrations, respectively, from `fv`-hypercolumns *`hladr.E`*. Function values from all `fv.object`s in the `fv`-hypercolumns *`hladr.E`* are legal, therefore, this aggregation is not truncated.

-   numeric-hypercolumns *`phenotype.G.value`* and *`phenotype.G.cumtrapz`*, the aggregated function values and aggregated cumulative trapezoidal integrations, respectively, from `fv`-hypercolumns *`phenotype.G`*. This aggregation is *truncated* at the minimum legal $r_\text{max}=240$ of the `fv.object`s in the `fv`-hypercolumn *`phenotype.G`*.

```{r}
(afv = out |>
  aggregate_fv(by = ~ patient_id, f_aggr_ = pmean))
```

User may specify an $r_\text{max}$ for truncation using the argument `rmax` of function `aggregate_fv()`.

```{r}
#| results: hide
out |>
  aggregate_fv(by = ~ patient_id, f_aggr_ = pmean, rmax = 220)
```

If the user-specified $r_\text{max}$ is greater than the minimum legal $r_\text{max}$, the user's $r_\text{max}$ will be ignored.

```{r}
#| results: hide
out |>
  aggregate_fv(by = ~ patient_id, f_aggr_ = pmean, rmax = 260)
```

🚧 **To be expanded to a section** Each of the numeric-hypercolumns contains tabulated values on the common grid of R. One "slice" of this grid may be extracted by

```{r}
afv$hladr.E.cumtrapz |> .slice(j = '50')
```

### Of numeric-hypercolumn, and numeric marks in `ppp`-hypercolumn

#### On `quantile`s

Function `aggregate_quantile()` aggregates the `quantile`s of the numeric-hypercolumns and the numeric marks in the `ppp`-hypercolumn.

In the following example, we have

-   numeric-hypercolumn *`phenotype.nncross.quantile`*, the aggregated `quantile`s of numeric-hypercolumn *`phenotype.nncross`*.
-   numeric-hypercolumn *`hladr.quantile`*, the aggregated `quantile`s of numeric mark *`hladr`* in `ppp`-hypercolumn.

```{r}
#| message: false
out |>
  aggregate_quantile(by = ~ patient_id, probs = seq.int(from = 0, to = 1, by = .1))
```

In this package, we include a `groupedHyperframe` example **`Ki67`** with a `ppp`-hypercolumn *`ppp.`* and a nested grouping structure *`~patientID/tissueID`*.

```{r}
data(Ki67, package = 'groupedHyperframe')
Ki67
```

We use function `aggregate_quantile()` to aggregate the `quantile`s of the numeric `marks`, i.e., the log-transformed Ki67 intensities, from each element of the `ppp`-hypercolumn *`ppp.`* at the second-lowest group `~patientID`.

```{r}
#| message: false
Ki67 |>
  aggregate_quantile(by = ~ patientID, probs = seq.int(from = .01, to = .99, by = .01)) |> 
  head()
```

Users are encouraged to learn more about the applications of the aggregated quantiles of **`Ki67`** data from package **`hyper.gam`** vignettes ([RPubs](https://rpubs.com/tingtingzhan/hyper_gam), [CRAN](https://CRAN.R-project.org/package=hyper.gam/vignettes/applications.html)), section *Quantile Index*, as well as from our peer-reviewed publications @Yi25; @Yi23a; @Yi23b.

#### On kernel densities

Function `aggregate_kerndens()` aggregates the kernel density of the numeric-hypercolumns and the numeric marks in the `ppp`-hypercolumn.

In the following example, we have

-   numeric-hypercolumn *`phenotype.nncross.kerndens`*, the aggregated kernel densities of numeric-hypercolumn *`phenotype.nncross`*.
-   numeric-hypercolumn *`hladr.kerndens`*, the aggregated kernel densities of numeric mark *`hladr`* in `ppp`-hypercolumn.

```{r}
#| message: false
(mdist = out$phenotype.nncross |> unlist() |> max())
out |> 
  aggregate_kerndens(by = ~ patient_id, from = 0, to = mdist)
```

# Appendix

We cover some minor and/or experimental features in the appendix, in order not to interrupt the main narrative of this vignette.

## On `groupedHyperframe`

### Nested Grouping Structure {#sec-nestedGroup}

Interaction terms using colon operator (`:`) are allowed in the grouping structure, such as `g1/g2a:g2b/g3a:g3b:g3c`. This feature is made possible because in R `language` (especially `formula`) object, the colon operator (`:`) has higher priority than the forward slash (`/`).

```{r}
#| code-fold: true
#| code-summary: "See for yourself"
quote(g1/g2a:g2b/g3a:g3b:g3c) |>
  as.list()
```

The helper function `get_nested()` breaks down a nested grouping structure by the forward slash (`/`).

```{r}
quote(g1/g2/g3) |>
  get_nested()
```

```{r}
quote(g1/g2a:g2b/g3a:g3b:g3c) |>
  get_nested()
```
The (nested) grouping structure `attr(., 'group')` of a `groupedHyperframe`

-   must denote the highest level using a `symbol` (e.g., `g1`)
-   may denote the lower levels using a `symbol` (e.g., `g2`) or an interaction-term (e.g., `g2a:g2b`)


## On `fv.object`

### Legal $r_\text{max}$ {#sec-legal_rmax}

In package **`spatstat.explore`** (up to version `r packageVersion('spatstat.explore')`), function `markcorr()` is the workhorse inside functions `Emark()`, `Vmark()` and `markvario()`. Function `markcorr()` relies on the un-exported workhorse function `spatstat.explore:::sewsmod()`, whose default `method = "density"` contains the calculation of the ratio of two kernel densities. Due to the floating-point precision of R, such density ratios may have exceptional/illegal returns of

-   `0`, from $0/\delta$, where $\delta\geq$ (approximately) `2.6e-324`
-   `NaN`, from $0/\varepsilon$, where $\varepsilon\leq$ (approximately) `2.5e-324`
-   `Inf`, from $\delta/0$, where $\delta\geq$ (approximately) `2.6e-324`

```{r}
#| code-fold: true
#| code-summary: "See for yourself"
0 / c(2.6e-324, 2.5e-324)
c(2.5e-324, 2.6e-324) / 0
```

Function `markcorr()` provides a default argument of parameter $r$, at which the mark correlation function $k_f(r)$ are evaluated. The S3 method dispatch `spatstat.explore::print.fv()` prints the *recommended range* and *available range* of the argument $r$.

```{r}
#| code-fold: true
#| code-summary: "Function spatstat.explore::print.fv() output"
spatstat.data::spruces |> 
  spatstat.explore::markcorr()
```

We may observe exceptional/illegal returns if we go beyond the recommended and/or available range. In the following example, we see that the mark correlation $k_f(r)$ (column *`iso`*) having exceptional/illegal values `0`, `NaN` and `Inf`.

```{r}
#| code-fold: true
#| code-summary: "Exceptions when parameter `r` out-of-range"
fv = spatstat.data::spruces |> 
  spatstat.explore::markcorr(r = 0:100)
fv |>
  spatstat.explore::as.data.frame.fv() |>
  utils::tail(n = 30L)
```

Function `lastLegal()` provides the *index* of the last legal function value in an `fv.object`.

```{r}
fv |>
  key1val.fv() |>
  lastLegal()
```

## On `ppp.object`

In this appendix section, we discuss a couple methods applicable to the `ppp.object`.

### Default $r_\text{max}$

The `S3` generic function `rmax_()` provides the default $r_\text{max}$ used in package **`spatstat.explore`** functions that returns an `fv.object`. The primary `S3` method dispatch `rmax_.ppp()` is an off-label use of the un-documented functions `spatstat.explore::rmax.rule()` and `spatstat.geom::handle.r.b.args()`.

| Function in package **`spatstat.explore`** | Call of `spatstat.explore::rmax.rule` | Default $r_\text{max}$ via `rmax_()` |
|------------------------|------------------------|------------------------|
| [`Emark` and `Vmark`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/Emark.html), [`markcorr`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/markcorr.html), [`markvario`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/markvario.html) | `rmax.rule(fun = 'K', ...)` | `rmax_(fun = 'K')` |
| [`Kcross`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/Kcross.html) , workhorse `Kest` and `Kmulti` | `rmax.rule(fun = 'K', ...)` | `rmax_(fun = 'K')` or `rmax_(fun = 'K', i, j)` |
| [`Gcross`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/Gcross.html), workhorse `Gest` and `Gmulti` | `rmax.rule(fun = 'G', ...)` | `rmax_(fun = 'G')` or `rmax_(fun = 'G', i, j)` |
| [`Jcross`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/Jcross.html), workhorse `Jest` and `Jmulti` | `rmax.rule(fun = 'J', ...)` | `rmax_(fun = 'J')` or `rmax_(fun = 'J', i, j)` |

: Default $r_\text{max}$ from functions in package **`spatstat.explore`** that return an `fv.object`

We demonstrate the use of `S3` method dispatch `rmax_.ppp()` using examples from package **`spatstat.data`**.

```{r}
#| code-fold: true
#| code-summary: "numeric-mark, rectangle-window, for function markcorr()"
spatstat.data::anemones |> rmax_(fun = 'K')
```

```{r}
#| code-fold: true
#| code-summary: "multitype-mark, rectangle-window, for function Gcross()"
spatstat.data::amacrine |> rmax_(fun = 'G')
spatstat.data::amacrine |> rmax_(fun = 'G', i = 'on', j = 'off')
spatstat.data::amacrine |> rmax_(fun = 'G', i = 'off', j = 'on')
```

```{r}
#| code-fold: true
#| code-summary: "multitype-mark, rectangle-window, for function Jcross()"
spatstat.data::paracou |> rmax_(fun = 'J')
spatstat.data::paracou |> rmax_(fun = 'J', i = 'adult', j = 'juvenile')
spatstat.data::paracou |> rmax_(fun = 'J', i = 'juvenile', j = 'adult')
```

Together with the additional `S3` method dispatches `rmax_.ppplist()` and `rmax_.hyperframe()`, user may obtain the default $r_\text{max}$ *before* applying the (potentially) very slow batch processes.

-   For functions `Emark_()`, `Vmark_()`, `markcorr_()`, and `markvario_()`,

```{r}
#| code-fold: false
#| code-summary: "fun = 'K', numeric (and multitype) mark"
s |> rmax_(fun = 'K')
```

-   For function `Kcross_()`,

```{r}
#| code-fold: false
#| code-summary: "fun = 'K', multitype mark"
s |> rmax_(fun = 'K')
s |> rmax_(fun = 'K', i = 'CK+.CD8-', j = 'CK-.CD8+')
s |> rmax_(fun = 'K', i = 'CK-.CD8+', j = 'CK+.CD8-')
```

-   For function `Gcross_()`,

```{r}
#| code-fold: false
#| code-summary: "fun = 'G', multitype mark"
s |> rmax_(fun = 'G')
s |> rmax_(fun = 'G', i = 'CK+.CD8-', j = 'CK-.CD8+')
s |> rmax_(fun = 'G', i = 'CK-.CD8+', j = 'CK+.CD8-')
```

-   For function `Jcross_()`,

```{r}
#| code-fold: false
#| code-summary: "fun = 'J', multitype"
s |> rmax_(fun = 'J')
s |> rmax_(fun = 'J', i = 'CK+.CD8-', j = 'CK-.CD8+')
s |> rmax_(fun = 'J', i = 'CK-.CD8+', j = 'CK+.CD8-')
```

### $k$-Means Clustering

The `S3` generic `.kmeans()` performs $k$-means clustering (workhorse function `stats::kmeans`).

```{r}
spatstat.data::shapley
```

The `S3` method dispatch `.kmeans.ppp()` performs $k$-means clustering, with paramters

-   `formula`, user-specified coordinate(s) and/or numeric `marks`;
-   `centers`, number of clusters, see `?stats::kmeans`
-   `clusterSize`, "expected" cluster size.

#### By coordinate(s) and/or `marks`

Example below shows a clustering based on the $x$- and $y$-coordinates, as well as the numeric mark *`Mag`*.

```{r}
km = spatstat.data::shapley |> .kmeans(formula = ~ x + y + Mag, centers = 3L)
km |> class()
```

Example below shows a clustering based on $x$-coordinate and *`Mag`*.

```{r}
km1 = spatstat.data::shapley |> .kmeans(formula = ~ x + Mag, centers = 3L)
km1 |> class()
```

Example below shows a clustering based on $x$- and $y$-coordinates only.

```{r}
km2 = spatstat.data::shapley |> .kmeans(formula = ~ x + y, centers = 3L)
km2 |> class()
```

#### By `clusterSize`

Example below shows a clustering specified by `clusterSize`.

```{r}
km3 = spatstat.data::shapley |> .kmeans(formula = ~ x + y, clusterSize = 1e3L)
km3 |> class()
km3$centers # 5 clusters needed
km3$cluster |> table()
```

#### Split by $k$-Means Clustering

The `S3` generic functoin `split_kmeans()` `split`s `ppp.object`, `listof` `ppp.object`s, and `hyperframe` by $k$-means clustering.

Note that many functions in package **`groupedHyperframe`** require a `'dataframe'` `markformat` for `ppp.object`s.

```{r}
data(flu, package = 'spatstat.data')
flu$pattern[[1L]] |> 
  spatstat.geom::markformat()
```

User may convert a `'vector'` `markformat` to `'dataframe'` using the syntactic sugar `` `mark_name<-` ``,

```{r}
flu$pattern[] = flu$pattern |> 
  lapply(FUN = `mark_name<-`, value = 'stain') # read ?flu carefully
flu$pattern[[1L]] |> 
  spatstat.geom::markformat()
```

##### Split a `ppp.object`

The `S3` method dispatch `split_kmeans.default()` splits a `ppp.object` by $k$-means clustering.

```{r}
flu$pattern[[1L]] |> split_kmeans(formula = ~ x + y, centers = 3L)
```

##### Split a `listof` `ppp.object`s

The `S3` method dispatch `split_kmeans.listof()` splits a `listof` `ppp.object`s by $k$-means clustering.

The returned object has `attributes`

-   `attr(.,'.id')`, indices of the `ppp.object`s before splitting.
-   `attr(.,'.cluster')`, indices of $k$-means clusters, nested in `.id`.

```{r}
flu$pattern[1:2] |> split_kmeans(formula = ~ x + y, centers = 3L) 
```

##### Split a `hyperframe` and/or `groupedHyperframe`

The `S3` method dispatch `split_kmeans.hyperframe()` splits a `hyperframe` and/or `groupedHyperframe` by $k$-means clustering of the *one-and-only-one* `ppp`-hypercolumn.

The returned object is a `groupedHyperframe` with grouping structure

-   `~.id/.cluster`, if the input is a `hyperframe`
-   `~ existing/grouping/structure/.cluster`, if the input is a `groupedHyperframe`. Note that the grouping level `.id` is **believed** to be equivalent to *the lowest level of existing grouping structure*.

```{r}
flu[1:2,] |> split_kmeans(formula = ~ x + y, centers = 3L)
```

### Pairwise Tjøstheim's Coefficient

The `S3` generic `pairwise_cor_spatial()` calculates the nonparametric, rank-based, Tjøstheim's correlation coefficients [@Tjostheim78; @Hubert82] in a pairwise-combination fashion, using the workhorse function `SpatialPack::cor.spatial()`. All `S3` method dispatches return a object of class `'pairwise_cor_spatial'`, which `inherits` from class `'dist'`.

The `S3` method dispatch `pairwise_cor_spatial.ppp()` finds the nonparametric Tjøstheim's correlation coefficients from the pairwise-combinations of all numeric marks of a `ppp.object`.

```{r}
(r = spatstat.data::finpines |> pairwise_cor_spatial())
```

The printing of `'pairwise_cor_spatial'` is taken care of by function `stats:::print.dist`.

The `S3` method dispatch `as.matrix.pairwise_cor_spatial()` returns a `matrix` of pairwise Tjøstheim's coefficient with `diag`onal values of 1. Note that this matrix is ***not*** a `cor`relation matrix, because Tjøstheim's correlation coefficient

-   is nonparametric, i.e., there is no definition of the corresponding `cov`ariance, standard deviation `sd`, nor the conversion `cov2cor` method;
-   does not provide a mathematical mechanism to ensure this matrix is [positive definite](https://en.wikipedia.org/wiki/Definite_matrix).

```{r}
r |> as.matrix()
```

```{=html}
<!--
# Appendix B: On `fv.object`

In this appendix section, we discuss a couple methods applicable to the `fv.object`.
-->
```

## What We Don't Do

Lastly, we keep note of the features that we plan **not** to support in package **`groupedHyperframe`**. Those features are technically tedious to implement, and/or have relatively easy workaround.

### `S4` method dispatch `merge`

The authors plan [**not**]{style="background-color: #FFFF00"} to implement an `S4` method dispatch `merge` (e.g., `S3` method dispatch `base::merge.data.frame()`) for `hyperframe` and/or `groupedHyperframe` classes, for several reasons.

-   There is not an `S3` method dispatch `merge.hyperframe` in package **`spatstat.geom`** (as of version `r packageVersion('spatstat.geom')`);
-   Should the authors decide to implement the `merge` functionality for `hyperframe` class, at least two `S4` method dispatches need to be written for `setMethod(f = merge, ...)`
    -   `signature = c(x = 'hyperframe', y = 'data.frame')`
    -   `signature = c(x = 'hyperframe', y = 'hyperframe')`
-   Should the authors decide to implement the `merge` functionality for `groupedHyperframe` class, at least three `S4` method dispatches need to be written for `setMethod(f = merge, ...)`
    -   `signature = c(x = 'groupedHyperframe', y = 'data.frame')`
    -   `signature = c(x = 'groupedHyperframe', y = 'hyperframe')`
    -   `signature = c(x = 'groupedHyperframe', y = 'groupedHyperframe')`, for which the authors need to consider the (potentially different) grouping structures of `x` and `y` inputs.

The authors suggest users do `base::merge.data.frame()` first, then do `as.groupedHyperframe()` as a workaround.

## Terms & Abbreviations {#sec-terms}

| Term / Abbreviation | Description |
|------------------------------------|------------------------------------|
| [`|>`](https://search.r-project.org/R/refmans/base/html/pipeOp.html) | Forward pipe operator introduced since R version 4.1.0 |
| [`:`](https://search.r-project.org/R/refmans/base/html/Colon.html) | Colon operator for [`interaction`](https://search.r-project.org/R/refmans/base/html/interaction.html) of [`factor`](https://search.r-project.org/R/refmans/base/html/factor.html)s |
| `/` | Forward slash operator to denote nested grouping structure `~ g1/.../gm` in grouped data frame [`groupedData`](https://search.r-project.org/CRAN/refmans/nlme/html/groupedData.html), from package **`nlme`** [@nlme] |
| [`attr`](https://search.r-project.org/R/refmans/base/html/attr.html), [`attributes`](https://search.r-project.org/R/refmans/base/html/attributes.html) | Attributes |
| CRAN, R | The Comprehensive R Archive Network, <https://cran.r-project.org> |
| [`cor`](https://search.r-project.org/R/refmans/stats/html/cor.html) | Correlation matrix |
| [`cor.spatial`](https://search.r-project.org/CRAN/refmans/SpatialPack/html/cor.spatial.html) | Tjøstheim's nonparametric correlation coefficient, from package **`SpatialPack`** [@SpatialPack] |
| [`cov`, `cov2cor`](https://search.r-project.org/R/refmans/stats/html/cor.html) | Variance-covariance matrix, and conversion to correlation matrix |
| [`data.frame`](https://search.r-project.org/R/refmans/base/html/data.frame.html) | Data frame |
| [`diag`](https://search.r-project.org/R/refmans/base/html/diag.html) | Matrix diagonals |
| [`dist`](https://search.r-project.org/R/refmans/stats/html/dist.html) | Distance matrix; to take advantage of `stats:::as.matrix.dist` |
| [`file.size`](https://search.r-project.org/R/refmans/base/html/file.info.html) | File size in bytes |
| (one-sided) [`formula`](https://search.r-project.org/R/refmans/stats/html/formula.html) | Formula, or [one-sided formula](https://search.r-project.org/R/refmans/stats/html/asOneSidedFormula.html) |
| [`fv`, `fv.object`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/fv.object.html), [`plot.fv`](https://search.r-project.org/CRAN/refmans/spatstat.explore/html/plot.fv.html) | (Plot of) function value table, from package **`spatstat.explore`** |
| [`groupedHyperframe`](https://CRAN.R-project.org/package=groupedHyperframe) | Grouped hyper data frame |
| `hypercolumns`, [`hyperframe`](https://search.r-project.org/CRAN/refmans/spatstat.geom/html/hyperframe.html) | (Hyper columns of) hyper data frame, from package **`spatstat.geom`** [@spatstat05] |
| [`inherits`](https://search.r-project.org/R/refmans/base/html/class.html) | Class inheritance |
| `kerndens` | Kernel density, `stats::density.default()$y` |
| [`Inf`](https://search.r-project.org/R/refmans/base/html/is.finite.html) | Positive infinity $\infty$ |
| [`kmeans`](https://search.r-project.org/R/refmans/stats/html/kmeans.html) | $k$-means clustering [@kmeans] |
| [`language`](https://search.r-project.org/R/refmans/base/html/is.language.html) | R language object |
| [`list`](https://search.r-project.org/R/refmans/base/html/list.html), [`listof`](https://search.r-project.org/R/refmans/stats/html/listof.html) | Lists of objects |
| [`.Machine`](https://search.r-project.org/R/refmans/base/html/zMachine.html) | Numerical characteristics of the machine R is running on, e.g., 32-bit integers and IEC 60559 floating-point (double precision) arithmetic |
| [`markformat`](https://search.r-project.org/CRAN/refmans/spatstat.geom/html/spatstat.geom-internal.html) | Storage mode of `marks` |
| [`marks`](https://search.r-project.org/CRAN/refmans/spatstat.geom/html/spatstat.geom-internal.html), [`marked`](https://search.r-project.org/CRAN/refmans/spatstat.geom/html/is.marked.html) | Marks of a point pattern |
| [`merge`](https://search.r-project.org/R/refmans/base/html/merge.html) | Merge two `data.frame`s |
| [`mc.cores`](https://search.r-project.org/R/refmans/parallel/html/mclapply.html) | Number of CPU cores to use for parallel computing |
| [`message`](https://search.r-project.org/R/refmans/base/html/message.html) | Diagnostic message printed in R console |
| [`multitype`](https://search.r-project.org/CRAN/refmans/spatstat.geom/html/spatstat.geom-internal.html) | Multitype spatial object |
| [`NaN`](https://search.r-project.org/R/refmans/base/html/is.finite.html) | Not-a-Number |
| [`object.size`](https://search.r-project.org/R/refmans/utils/html/object.size.html) | Memory allocation |
| `pmean`, `pmedian` | Point-wise mean and median, from package `groupedHyperframe` |
| [`pmax`, `pmin`](https://search.r-project.org/R/refmans/base/html/Extremes.html) | Point-wise maxima and minima |
| [`ppp`](https://search.r-project.org/CRAN/refmans/spatstat.geom/html/ppp.html), [`ppp.object`](https://search.r-project.org/CRAN/refmans/spatstat.geom/html/ppp.object.html) | (Marked) point pattern |
| [`quantile`](https://search.r-project.org/R/refmans/stats/html/quantile.html) | Quantile |
| [`save`](https://search.r-project.org/R/refmans/base/html/save.html), [`saveRDS`](https://search.r-project.org/R/refmans/base/html/readRDS.html), `xz` | Save with [`xz`](https://en.wikipedia.org/wiki/XZ_Utils) compression |
| `S3`, `generic`, [`methods`](https://search.r-project.org/R/refmans/utils/html/methods.html) | `S3` object oriented system, [`UseMethod`](https://search.r-project.org/R/refmans/base/html/UseMethod.html); [`getS3method`](https://search.r-project.org/R/refmans/utils/html/getS3method.html); <https://adv-r.hadley.nz/s3.html> |
| `S4`, `generic`, `methods` | `S4` object oriented system, [`isS4`](https://search.r-project.org/R/refmans/base/html/isS4.html); [`setClass`](https://search.r-project.org/R/refmans/methods/html/setClass.html); [`setMethod`](https://search.r-project.org/R/refmans/methods/html/setMethod.html); [`getMethod`](https://search.r-project.org/R/refmans/methods/html/getMethod.html); <https://adv-r.hadley.nz/s4.html> |
| [`sd`](https://search.r-project.org/R/refmans/stats/html/sd.html) | Standard deviation |
| [`search`](https://search.r-project.org/R/refmans/base/html/search.html) | Search path |
| [`Surv`](https://search.r-project.org/CRAN/refmans/survival/html/Surv.html) | Survival, i.e., time-to-event, object, from package **`survival`** [@survival] |
| [`symbol`, `name`](https://search.r-project.org/R/refmans/base/html/name.html) | R object name/symbol |
| [`trapz`, `cumtrapz`](https://search.r-project.org/CRAN/refmans/pracma/html/trapz.html) | (Cumulative) [trapezoidal integration](https://en.wikipedia.org/wiki/Trapezoidal_rule), from package **`pracma`** [@pracma] |
| [`vector`](https://search.r-project.org/R/refmans/base/html/vector.html) | Vector |


# References

::: {#refs}
:::
