---
title: "Grouped Hyperframe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```


# Introduction

This vignette for package **`groupedHyperframe`** documents the creation of `groupedHyperframe` object, the batch processes defined for a `groupedHyperframe`, and aggregations over multi-level grouping structure.

Package **`groupedHyperframe`** requires the development versions of **`spatstat`** family of packages.
```{r eval = FALSE}
devtools::install_github('spatstat/spatstat'); packageDate('spatstat')
devtools::install_github('spatstat/spatstat.data'); packageDate('spatstat.data')
devtools::install_github('spatstat/spatstat.explore'); packageDate('spatstat.explore')
devtools::install_github('spatstat/spatstat.geom'); packageDate('spatstat.geom')
devtools::install_github('spatstat/spatstat.linnet'); packageDate('spatstat.linnet')
devtools::install_github('spatstat/spatstat.model'); packageDate('spatstat.model')
devtools::install_github('spatstat/spatstat.random'); packageDate('spatstat.random')
devtools::install_github('spatstat/spatstat.sparse'); packageDate('spatstat.sparse')
devtools::install_github('spatstat/spatstat.univar'); packageDate('spatstat.univar')
devtools::install_github('spatstat/spatstat.utils'); packageDate('spatstat.utils')
```

A development version of package **`groupedHyperframe`** is hosted on [Github](https://github.com/tingtingzhan/groupedHyperframe).

```{r eval = FALSE}
devtools::install_github('tingtingzhan/groupedHyperframe'); packageDate('groupedHyperframe')
```

Examples in this vignette require the **`survival`** package.

```{r setup}
library(groupedHyperframe)
library(survival) # to help ?spatstat.geom::hyperframe understand ?survival::Surv object
```


# `groupedHyperframe` Class

The S3 class `groupedHyperframe` inherits from `spatstat.geom::hyperframe`, in a similar fashion as `nlme::groupedData` inherits from `base::data.frame`.

A `groupedHyperframe` object, in addition to `spatstat.geom::hyperframe` object, has attribute(s)

* `attr(., 'group')`, a `formula` to specify the grouping structure



## With one-and-only-one `spatstat.geom::ppp`-hypercolumn

Function `grouped_ppp()` creates a `groupedHyperframe` with one-and-only-one marked point pattern (`spatstat.geom::ppp`) hypercolumn.  In the following example, the argument `formula` specifies

* the marks, e.g., `numeric` mark *`hladr`* and `spatstat.geom::is.multitype` mark *`phenotype`*, on the left-hand-side
* additional predictors and/or endpoints, e.g., *`OS`*, *`gender`* and *`age`*, before the `|` separator on the right-hand-side
* grouping structure, e.g., *`image_id`* nested in *`patient_id`*, after the `|` separator on the right-hand-side. Users may see `?nlme::groupedData` or `?nlme::lme` for a thorough discussion on the syntax of nested grouping structure `~g1/.../gm`. 

```{r}
(s = grouped_ppp(formula = hladr + phenotype ~ OS + gender + age | patient_id/image_id, 
                 data = wrobel_lung))
```

Function `grouped_ppp()` has parameter `coords` which specifies the column name of $x$- and $y$-coordinates in the input `data`. Default `coords = ~ x + y` indicates the use of `data$x` and `data$y` for $x$- and $y$-coordinates, respectively.  Users may use `coords = FALSE` for data without $x$- and $y$-coordinates. In this case, the coordinates are filled with randomly generated numbers, and the returned `groupedHyperframe` has a `pseudo.ppp`-hypercolumn.

```{r}
(s_a = grouped_ppp(Ki67 ~ Surv(recfreesurv_mon, recurrence) + race + age | patientID/tissueID, 
                  data = Ki67, coords = FALSE))
```


# Batch Processes on `ppp`-Hypercolumn 

In this section, we outline the batch processes of spatial point pattern analyses applicable to the one-and-only-one `ppp`-hypercolumn of a `hyperframe`.  

Note that these spatial point pattern analyses should **not** be applied to a `pseudo.ppp`-hypercolumn, as the $x$- and $y$-coordinates are randomly generated psuedo numbers.


Batch process(es) that return a `spatstat.explore::fv.object`-hypercolumn include

```{r echo = FALSE, results = 'asis'}
data.frame(
  Function = c('`Emark_()`', '`Vmark_()`', '`markcorr_()`', '`markvario_()`', '`Gcross_()`', '`Kcross_()`', '`Jcross_()`'),
  Workhorse = c('`spatstat.explore::Emark`', '`spatstat.explore::Vmark`', '`spatstat.explore::markcorr`', '`spatstat.explore::markvario`', '`spatstat.explore::Gcross`', '`spatstat.explore::Kcross`', '`spatstat.explore::Jcross`'),
  Applicable = c('`numeric` marks (e.g., *`hladr`*) in `ppp`-hypercolumn', '`numeric` marks', '`numeric` marks', '`numeric` marks', '`multitype` marks (e.g., *`phenotype`*)', '`multitype` marks', '`multitype` marks')
) |> kable(caption = 'Batch processes that return `fv`-hypercolumn')
```


Batch process(es) that return a `numeric`-hypercolumn include

```{r echo = FALSE, results = 'asis'}
data.frame(
  Function = c('`nncross_()`'),
  Workhorse = c('`spatstat.geom::nncross.ppp(., what = \'dist\')`'),
  Applicable = c('`multitype` marks (e.g., *`phenotype`*)')
) |> kable(caption = 'Batch processes that return `numeric`-hypercolumn')
```


Following example shows that the batch process(es) may be applied to a `groupedHyperframe` in a pipeline (`|>`).

```{r}
r = seq.int(from = 0, to = 250, by = 10)
out = s |>
  Emark_(r = r, correction = 'best') |> # slow
  # Vmark_(r = r, correction = 'best') |> # slow
  # markcorr_(r = r, correction = 'best') |> # slow
  # markvario_(r = r, correction = 'best') |> # slow
  Gcross_(i = 'CK+.CD8-', j = 'CK-.CD8+', r = r, correction = 'best') |> # fast
  # Kcross_(i = 'CK+.CD8-', j = 'CK-.CD8+', r = r, correction = 'best') |> # fast
  nncross_(i = 'CK+.CD8-', j = 'CK-.CD8+', correction = 'best') # fast
```

The returned `groupedHyperframe` has new hypercolumns

* `fv`-hypercolumn *`hladr.E`*, created by function `Emark()` on `numeric` mark *`hladr`*
* `fv`-hypercolumn *`phenotype.G`*, created by function `Gcross_()` on `multitype` mark *`phenotype`*
* `numeric`-hypercolumn *`phenotype.nncross`*, created by function `nncross_()` on `multitype` mark *`phenotype`*

```{r}
out
```

# Aggregation Over Nested Grouping Structure

When nested grouping structure `~g1/g2/.../gm` is present, we may aggregate over the

* `fv`-hypercolumn(s)
* `numeric`-hypercolumn(s)
* `numeric` marks in the `ppp`-hypercolumn

by either one of the grouping levels `~g1`, `~g2`, ..., or `~gm`. If the lowest grouping `~gm` is specified, then no aggregation is performed.

The returned object of various aggregation functions, `aggregate_fv()`, `aggregate_quantile()` and `aggregate_kerndens()`, is `base::data.frame` instead of `spatstat.geom::hyperframe`.  This is because the aggregated results are stored in `matrix`-columns, while `spatstat.geom::hyperframe` does not support `matrix`-column.


## Aggregation of `fv`-hypercolumn(s)

Function `aggregate_fv()` aggregates 

* the **function values**, i.e., the black-solid-curve of `spatstat.explore::plot.fv`. In the following example, we have
  + `matrix`-column *`hladr.E.value`*, aggregated function value from `fv`-hypercolumn *`hladr.E`*
  + `matrix`-column *`phenotype.G.value`*, aggregated function value from `fv`-hypercolumn *`phenotype.G`*
* the **cumulative trapezoid area** under the black-solid-curve, via function `pracma::cumtrapz`.  In the following example, we have
  + `matrix`-column *`hladr.E.cumarea`*, aggregated cumulative trapezoid area from `fv`-hypercolumn *`hladr.E`*
  + `matrix`-column *`phenotype.G.cumarea`*, aggregated cumulative trapezoid area from `fv`-hypercolumn *`phenotype.G`*

```{r}
afv = out |>
  aggregate_fv(by = ~ patient_id, f_aggr_ = 'mean')
nrow(afv) # number of patients
names(afv)
dim(afv$hladr.E.cumarea) # N(patient) by length(r)
```

## Aggregation of `numeric`-hypercolumn(s) and `numeric` mark(s) in `ppp`-hypercolumn

Function `aggregate_quantile()` aggregates 

* the `stats::quantile` of the `numeric`-hypercolumn(s).  In the following example, we have
  + `matrix`-column *`phenotype.nncross.quantile`*, aggregated `quantile` of `numeric`-hypercolumn *`phenotype.nncross`*
* the `quantile` of the `numeric` mark(s) in the `ppp`-hypercolumn.  In the following example, we have
  + `matrix`-column *`hladr.quantile`*, aggregated `quantile` of `numeric` mark *`hladr`* in `ppp`-hypercolumn

```{r}
q = out |>
  aggregate_quantile(by = ~ patient_id, probs = seq.int(from = 0, to = 1, by = .1))
nrow(q)
names(q)
dim(q$phenotype.nncross.quantile)
dim(q$hladr.quantile)
```

Function `aggregate_kerndens()` aggregates 

* the kernel density `stats::density.default()$y` from the `numeric`-hypercolumn(s).  In the following example, we have
  + `matrix`-column *`phenotype.nncross.kerndens`*, aggregated kernel density of `numeric`-hypercolumn *`phenotype.nncross`*
* the kernel density of the `numeric` mark(s) in the `ppp`-hypercolumn.  In the following example, we have
  + `matrix`-column *`hladr.kerndens`*, aggregated kernel density of `numeric` mark *`hladr`* in `ppp`-hypercolumn

```{r}
(mdist = out$phenotype.nncross |> unlist() |> max())
d = out |> 
  aggregate_kerndens(by = ~ patient_id, from = 0, to = mdist)
nrow(d)
names(d)
dim(d$phenotype.nncross.kerndens)
```