---
title: "Split by $k$-Means Clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{split_kmeans}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Introduction

This vignette of package **`groupedHyperframe`** ([CRAN](https://cran.r-project.org/package=groupedHyperframe), [Github](https://github.com/tingtingzhan/groupedHyperframe), [RPubs](https://rpubs.com/tingtingzhan/groupedHyperframe)) documents the $k$-means clustering of `ppp.object`, and the `split`ting of `ppp.object`, `listof`, and `hyperframe` by $k$-means clustering.

## Note to Users

Examples in this vignette require that the `search` path has

```{r setup}
library(groupedHyperframe)
library(spatstat.data)
library(spatstat.geom)
```

## Terms and Abbreviations

```{r echo = FALSE, results = 'asis'}
c(
  '', 'Forward pipe operator', '`?base::pipeOp` introduced in `R` 4.1.0', 
  '`attr`', 'Attributes', '`base::attr`; `base::attributes`',
  '`CRAN`, `R`', 'The Comprehensive R Archive Network', 'https://cran.r-project.org',
  '`data.frame`', 'Data frame', '`base::data.frame`',
  '`formula`', 'Formula', '`stats::formula`',
  '`groupedHyperframe`', 'Grouped hyper data frame', 'groupedHyperframe::as.groupedHyperframe',
  '`hypercolumns`, `hyperframe`', '(Hyper columns of) hyper data frame', '`spatstat.geom::hyperframe`',
  '`inherits`', 'Class inheritance', '`base::inherits`',
  '`kmeans`', '$k$-means clustering', '`stats::kmeans`',
  '`listof`', 'Lists of objects', '`?stats::listof`',
  '`markformat`', 'Storage mode of `marks`', '`spatstat.geom::markformat`',
  '`marks`', 'Marks of a point pattern', '`spatstat.geom::marks`',
  '`mc.cores`', 'Number of CPU cores to use', '`parallel::mclapply`; `parallel::detectCores`',
  '`multitype`', 'Multitype object', '`spatstat.geom::is.multitype`',
  '`object.size`', 'Memory allocation', '`utils::object.size`',
  '`ppp`, `ppp.object`', '(Marked) point pattern', '`spatstat.geom::ppp.object`',
  '`save`, `xz`', 'Save with `xz` compression', '`base::save(., compress = \'xz\')`; `base::saveRDS(., compress = \'xz\')`; https://en.wikipedia.org/wiki/XZ_Utils', 
  '`S3`, `generic`, `methods`', '`S3` object oriented system',  '`base::UseMethod`; `utils::methods`; `utils::getS3method`; https://adv-r.hadley.nz/s3.html',
  '`search`', 'Search path', '`base::search`'
) |>
  matrix(nrow = 3L, dimnames = list(c('Term / Abbreviation', 'Description', 'Reference'), NULL)) |>
  t.default() |>
  as.data.frame.matrix() |> 
  kable(format = 'html') 
# ?knitr::kable
# default: `|` shown as &...
# format = 'html': `>` shown as &..
```

## Acknowledgement

This work is supported by NCI R01CA222847 ([I. Chervoneva](https://orcid.org/0000-0002-9104-4505), [T. Zhan](https://orcid.org/0000-0001-9971-4844), and [H. Rui](https://orcid.org/0000-0002-8778-261X)) and R01CA253977 (H. Rui and I. Chervoneva).


# $k$-Means Clustering

S3 generic `.kmeans()` has the workhorse function `stats::kmeans`.

## $k$-means clustering on `ppp.object`

Data set `spatstat.data::shapley`.
```{r fig.width=5}
shapley |> plot()
```

The `S3` method dispatch `.kmeans.ppp()` performs $k$-means clustering based on one-or-more user-specified `numeric` coordinates and/or `marks`.

Example below shows a clustering based on the $x$- and $y$-coordinates, as well as the `numeric`-`mark` *`Mag`*.
```{r}
km = shapley |> .kmeans(formula = ~ x + y + Mag, centers = 3L)
km |> class()
```

Example below shows a clustering based on $x$-coordinate and *`Mag`*.
```{r}
km1 = shapley |> .kmeans(formula = ~ x + Mag, centers = 3L)
km1 |> class()
```

Example below shows a clustering based on $x$- and $y$-coordinates only.
```{r}
km2 = shapley |> .kmeans(formula = ~ x + y, centers = 3L)
km2 |> class()
```

# Split by $k$-Means Clustering

Many functions in package **`groupedHyperframe`** requires a `'dataframe'` `markformat` for `ppp.object`s. 

```{r}
flu$pattern[[1L]] |> markformat()
```

User may convert a `'vector'` `markformat` to `'dataframe'` using the syntactic sugar `` `mark_name<-` ``,

```{r}
flu$pattern[] = flu$pattern |> 
  lapply(FUN = `mark_name<-`, value = 'stain') # read ?flu carefully
flu$pattern[[1L]] |> markformat()
```

## Split `ppp.object`

The `S3` method dispatch `split_kmeans.default()` splits a `ppp.object` (more to come in future) by $k$-means clustering.

```{r}
flu$pattern[[1L]] |> split_kmeans(formula = ~ x + y, centers = 3L)
```

## Split a `listof` `ppp.object`s 

The `S3` method dispatch `split_kmeans.listof()` splits a `listof` `ppp.object`s by $k$-means clustering.

The returned object has `attributes`

* `attr(.,'.id')`, indices of the `ppp.object`s before splitting.
* `attr(.,'.cluster')`, indices of $k$-means clustering, nested in `.id`.

```{r}
flu$pattern[1:2] |> split_kmeans(formula = ~ x + y, centers = 3L) 
```

## Split a `hyperframe` and/or `groupedHyperframe`

The `S3` method dispatch `split_kmeans.hyperframe()` splits a `hyperframe` and/or `groupedHyperframe` by $k$-means clustering of the *one-and-only-one* `ppp`-`hypercolumn`.  

The returned object is a `groupedHyperframe` with grouping structure

* `~.id/.cluster`, if the input is a `hyperframe`
* `~ existing/grouping/structure/.cluster`, if the input is a `groupedHyperframe`. Note that the grouping level `.id` is **believed** to be equivalent to *the lowest level of existing grouping structure*.

```{r}
flu[1:2,] |> split_kmeans(formula = ~ x + y, centers = 3L)
```
